 

### 2021-08-26 vertical-thermal-review.Rmd
Marielle's citations and ET imag

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed an average of $1.5^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; IPBES report], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

@@

![**Figure 5.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests. Panel (a) is from (Kunert et al. 2017). I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/ecosystem partitioning_updated.png)

@@

Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019]. This buffering is influenced by hydrology, as wetter sites and seasons allow for denser, and more buffered canopies  [@davisMicroclimaticBufferingForests2019]. Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015]. The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging (Laurance et al. 2006; Miller et al. 2004), although fire tends to predominantly affect smaller trees (Brando et al. 2019). Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]

In general, disturbances (e.g., drought, fire, fragmentation, and logging) reduce canopy cover, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020]. This shift to hotter and drier microclimates often makes forests more susceptible to further disturbances, for example, increasing fire risk (Brando et al. 2014; Aragao et al. 2018). Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization'; Silverio et al. 2013) and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].

>>>>

**Forest responses and feedbacks to climate change will have a critical influence on the future of Earth's climate.**
Globally, temperatures have risen $1.2^\circ$C since 1900 [@rohdeBerkeleyEarthLand2020] and are expected to reach or even exceed an average of $1.5^\circ$C in the next 20 years [@ipccClimateChangeWidespread2021], accompanied by increasing severity and frequency of heat waves [@meehlMoreIntenseMore2004; @ipccClimateChangeWidespread2021] and hotter droughts [@trenberthGlobalWarmingChanges2014b].
These changes are expected to have profound effects on tree metabolism and forest ecosystem function [@breshearsUnderappreciatedPlantVulnerabilities2021; @portnerScientificOutcomeIPBESIPCC2021], by altering rates of photosynthesis and respiration [@breshearsUnderappreciatedPlantVulnerabilities2021; @corlettImpactsWarmingTropical2011; @scafaroResponsesLeafRespiration2021], causing foliar damage during heat waves [@osullivanThermalLimitsLeaf2017; @corlettImpactsWarmingTropical2011], and reducing growth and elevating mortality during drought [@rollinsonClimateSensitivityUnderstory2020; @breshearsUnderappreciatedPlantVulnerabilities2021; @mcdowellPervasiveShiftsForest2020]. Collectively, these impacts may lead to reduction of forest carbon sequestration, as has already occurred in some forests [@oishiWarmerTemperaturesReduce2018; @sullivanLongtermThermalSensitivity2020]. The resulting feedbacks to the climate system, carbon storage, and changes in albedo and hydrology will in turn impact the the future trajectory of climate change [@bonan_ecological_2016], yet the degree of the impact remains uncertain [@krauseLargeUncertaintyCarbon2018; @friedlingsteinClimateCarbonCycle2006].

@@

![**Figure 5.** Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests. Panel (a) is from (Kunert et al. 2017). I'd like to create a parallel figure for C fluxes.](schematics/ecosystem vertical partitioning/raw from Nobby/Fig_5_ET.png)

@@

Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019]. This buffering is influenced by hydrology, as wetter sites and seasons allow for denser, and more buffered canopies  [@davisMicroclimaticBufferingForests2019]. Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015]. The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging [[@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019]. Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]

In general, disturbances (e.g., drought, fire, fragmentation, and logging) reduce canopy cover, which in turn increases incident radiation levels and temperatures within the canopy [@starkReframingTropicalSavannization2020]. This shift to hotter and drier microclimates often makes forests more susceptible to further disturbances, for example, increasing fire risk [@brandoAbruptIncreasesAmazonian2014; @aragao21stCenturyDroughtrelated2018]. Severe degradation impacts can cause dramatic ecological state changes (e.g., the transition from forest to more open, savanna-like vegetation in tropical forest regions through 'savannization') and non-linear threshold responses in energy balance and associated microclimates, with implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020].
